\begin{longtable}{c|c|c}
    \caption{Progress}
    \label{tab:plan}
    
    \hline
    \multicolumn{3}{c}{\textbf{Legend}} \\
    \hline
    $\boxed{\checkmark}$ & Completed & \\
    \hline
    $\boxed{\times}$ & Not started & \\
    \hline
    $\boxed{+}$ & Started; needs work & \\
    \hline
    $\boxed{-}$ & Remove or scale down & \\
    
    \hline
    \multicolumn{3}{c}{\textbf{Data}} \\
    \hline
    $\boxed{\checkmark}$ & MNIST & \\ 
    \hline
    \multirow{3}{*}{$\boxed{-}$} & \multirow{3}{*}{Microscopy data} & 
    Requires too much context to explain. \\
    & & It's interesting to see how DeePWAK finds structure in a low quality \\
    & & data set but it's still a low quality data set. \\
    
    \hline
    \multicolumn{3}{c}{\textbf{Outer models}} \\
    \hline
    $\boxed{\checkmark}$ & Encoder & \\
    \hline
    $\boxed{\checkmark}$ & Classifier & \\
    \hline
    $\boxed{\checkmark}$ & Decoder & \\

    \hline
    \multicolumn{3}{c}{\textbf{Inner models}} \\
    \hline
    \multirow{2}{*}{$\boxed{\checkmark}$} & \multirow{2}{*}{SAE} &
    A minimal SAE implementation is available at \\
    & & \url{https://github.com/kewiechecki/SAE} \\
    \hline
    $\boxed{\checkmark}$ & PSAE & see above \\
    \hline
    \multirow{4}{*}{$\boxed{-}$} &  \multirow{4}{*}{DeePWAK} &
    Might be needlessly complicating things. \\
    & & Not straightforward to interpret embed \& unembed. \\
    & & If I want to look at importance of nonlinearity \\
    & & it makes more sense to look at SAE with no $\relu$. \\
    \hline
    \multirow{2}{*}{$\boxed{-}$} & \multirow{2}{*}{DeePWAKBlock} &
    This can be relegated to a separate paper. \\
    & & It doesn't add much for the model I'm working with. \\
    \hline

    \multicolumn{3}{c}{\textbf{Inner model methods}} \\
    \hline
    $\boxed{\checkmark}$ & Encoding & \\
    \hline
    $\boxed{\checkmark}$ & Decoding & \\
    \hline
    $\boxed{\checkmark}$ & Clustering & \\
    \hline
    $\boxed{\checkmark}$ & Cluster centroids & \\
    \hline
    
    \multicolumn{3}{c}{\textbf{Training}} \\
    \hline
    $\boxed{\times}$ & Checkpoints & Save models at different stages of training \\
    \hline

    \subsection{Visualizations}

$\boxed{\checkmark}$ & Loss} Figs. \ref{fig:lossouter}, \ref{fig:lossinner &
$\boxed{\times}$ & Visualize cluster centroids throughout training &

\multicolumn{3}{c}{\textbf{Heatmaps}} \\

$\boxed{+}$ & How to read &
\S \ref{sec:heatmaps}
$\boxed{\checkmark}$ & Embeddings &
Fig. \ref{fig:E_MNIST}
$\boxed{\checkmark}$ & Clusters &
Fig. \ref{fig:K_MNIST}
\multicolumn{3}{c}{\textbf{Hypergeometric test}} \\

$\boxed{\times}$ & How to read dot plot &
$\boxed{\checkmark}$ & Dot plot &
Fig. \ref{fig:hyperMNIST}
$\boxed{\times}$ & Remove outline scale from dot plot &
This is confusing and doesn't add useful information.

\subsection{Distribution}
$\boxed{+}$ & Git repo &
Code is divided between
\url{https://github.com/kewiechecki/DeePWAK} and
\url{https://github.com/kewiechecki/SAE}.
These need to be unified.

\multicolumn{3}{c}{\textbf{Documentation}} \\
$\boxed{+}$ & Model documentation &
$\boxed{+}$ & Training documentation &
$\boxed{\times}$ & Figure documentation &
$\boxed{\times}$ & Julia modules &

\subsection{Publication}
$\boxed{\checkmark}$ & \LaTeX template &
$\boxed{\checkmark}$ & Share on Overleaf &
\hyperlink{https://www.overleaf.com/read/mcvdvdfyfthn#fb5a8e}{view}

\multicolumn{3}{c}{\textbf{Illustrative figures}} \\
$\boxed{\times}$ & Outer models &
$\boxed{\checkmark}$ & DeePWAK flowchart &
Fig. \ref{fig:deepwak}
$\boxed{\times}$ & PWAK example &
$\boxed{+}$ & Diffusion example &
Take out DEWAKSS example. Substitute example from actual data.
$\boxed{+}$ & Figure legends &

\multicolumn{3}{c}{\textbf{Algorithms}} \\
$\boxed{-}$ & noise2self &
$\boxed{-}$ & PWAK &

\subsection{Supplementary Publications}
$\boxed{\times}$ & Theoretical foundations &
Material cut from \S \ref{sec:2}. Probably post to LessWrong.
$\boxed{+}$ & DeePWAK &
Split out all microscopy and multihead stuff to
\url{https://github.com/kewiechecki/DeePWAK}
%\url{https://drive.google.com/file/d/1geDArkvUQVOp79dF9knNmO3iP-1BYGOD/view?usp=drive_link}.

    
\end{longtable}